Leveraging genomics to detect trait selection from within and between-species variations

### 2023-09-13 manuscript/main.tex
M. Bastian, T. Gaboriau and N. Salamin edits

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}
\usepackage{caption}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,intitle=true,style=nature,date=year]{biblatex}
\addbibresource{references.bib}

@@

% -- Defining colors:
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}% Definig a custom style:
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    basicstyle=\ttfamily\scriptsize\bfseries,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=t,
    keepspaces=true,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}% -- Setting up the custom style:
\lstset{style=mystyle}

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\Trait_{\Root}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{G}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{G}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\MatrixGenetic}{\BiDimArray{G}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

% Including genomics, Inclusion of genomics, Linking genomics,
\title{Leveraging genomics to detect trait selection from within and between species variations}
\rhead{\scshape Trait selection from within and between species variations}

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M. {Bastian}$^{2}$, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
$^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}

\begin{document}

@@

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between species variation.
    Using within species variation, however, methods do not integrate the changes at the macro-evolutionary scale.
    Conversely, using between species variation, current methods usually discard within species variation, thus not accounting for processes at the micro-evolutionary scale.
    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
    As such, it does not require estimation of population size nor of time of speciation for normalization.
    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
\end{abstract}

\keywords{Quantitative genetics \and Trait evolution \and Selection }

@@

Determining whether a particular trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits and sequences.
Such variation can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these scales require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\cite{lynch_genetics_1998, walsh_evolution_2018}.

First, leveraging individual variations within the same species, Genome Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated to a given trait), pleiotropic (many traits associated to a given locus), additive and mainly under stabilizing selection~\cite{simons_population_2018, sella_thinking_2019}.
Additionally, at the macro-evolutionary scale, by contrasting both trait and genetic differentiation across populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to quantify selection acting on a trait~\cite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of expression of genes in different populations~\cite{whitehead_neutral_2006, gilad_natural_2006}.
However, dominant selective regime acting on gene expression is still controversial and neutral evolution is still debated~\cite{signor_evolution_2018, price_detecting_2022}.
For a pair of species, change in traits accumulate linearly with time from divergence and proportionally to the trait variance~\cite{lande_genetic_1980, turelli_heritable_1984}.
Gene expression exhibit such accumulation and divergence in expression accumulate faster for gene with large within species variation~\cite{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\cite{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, at the macro-evolutionary scale, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\cite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\cite{hansen_translating_1996}.
As a result, the mean trait value along the whole phylogeny can be modeled as a Brownian process branching at every node of the tree, allowing theoretically to test for neutrality~\cite{hansen_translating_1996, harmon_phylogenetic_2018}.
A deviation from the Brownian process, typically an Ornstein-Uhlenbeck process, is interpreted as a signature of stabilizing selection~\cite{catalan_drift_2019}.
Alternatively, a rapid shift in mean trait value along a branch is interpreted as a signature of directional selection.
However, studies have shown that such comparative approaches are subject to different biases~\cite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\cite{hansen_translating_1996}.
In other words, a best fit for a Brownian process is not necessarily a proof of the neutral model.
Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\cite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Altogether, methods at the phylogenetic scale easily misclassify selection and the mean trait value is taken either as the trait for a single individual in the species or as the average trait across several individuals, leaving out the variance in traits between individuals.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macro evolution, comparative methods at the phylogenetic scale have acknowledged the importance to model within species variation on top of changes in mean trait value~\cite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020}.
However, within species variation is used to describe technical errors, or the ratio of between to within variation is estimated for many traits, and is finally compared to the average over all traits to seek deviation from this average~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
Determining the evolutionary regime of a quantitative trait through articulation of the variance between and within species is the goal is this study, while setting threshold for neutral, stabilizing and diversifying selection.
In this study, we propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimate of population size (within species) nor of time of speciation (between species).
For population geneticist, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
For a phylogeneticist, our study can be seen as an extension of the EVE model~\cite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the man of variances

\section*{Materials and Methods}\label{sec:materials-and-methods}
\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}
\subsubsection*{Mutation-drift equilibrium for a trait}

For a given trait, its genetic architecture is defined by the number of loci encoding the trait ($\NbrLoci$) and the random effect of a mutation on the trait ($a$).
New mutations are thus generating variance for such a trait, where the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$ as:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

Mutations supply new variants, while genetic random drift depletes standing variation.
As shown in \textcite{lynch_mutation_1998}, if the trait is neutral, at equilibrium between mutation and drift, the genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$) as:
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating divided by the size of the region.
Any of the segregating mutations will eventually reach fixation or extinction due to genetic drift, such that the number of segregating mutations is also a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$) as:
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

Altogether, we define $\RateWhithin$ as the ratio of genetic variance of the trait ($\VarGenetic$) over nucleotide diversity of any neutral genomic region of interest ($\pi$).
This ratio allows to discard the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

Additionally, the genetic variance is equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by heritability ($\Heritability$), also giving $\RateWhithin$ as a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

\subsubsection*{Phylogenetic evolution of a trait}

At generation $\Time$, we denote $\MeanTrait$ as the mean value of the trait over the individuals in the species.
If the trait is neutral and encoded by many loci, $\MeanTrait$ evolves as Brownian process~\cite{hansen_translating_1996}.
After the species has evolved for $\Time$ generations, the variance of $\MeanTrait$, $\VarPhy$, is given by \textcite{hansen_translating_1996} as:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any neutral genomic region of interest, some will eventually reach fixation due to genetic drift, resulting in a substitution at the species level.
For a neutral mutation, its probability of fixation ($\pfix$) is $1/2\Ne$, as shown in \textcite{kimura_probability_1962}.
As reviewed in \textcite{mccandlish_modeling_2014}, we can thus derive the substitution rate per generation, denoted $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each of these newly arisen mutations $\pfix$, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate with which new neutral alleles are fixed in a genomic region of interest equals the rate with which new mutations arise per generation for the same genomic region~\cite{kimura_evolutionary_1968}.

>>>>

%! BibTeX Compiler = biber
%TC:ignore
\documentclass{article}
\usepackage{caption}
\usepackage{censor}
\usepackage{xcolor, colortbl}
\definecolor{BLUELINK}{HTML}{0645AD}
\definecolor{DARKBLUELINK}{HTML}{0B0080}
\definecolor{LIGHTGREY}{gray}{0.9}
\PassOptionsToPackage{hyphens}{url}
\usepackage[colorlinks=false]{hyperref}
% for linking between references, figures, TOC, etc in the pdf document
\hypersetup{colorlinks,
    linkcolor=DARKBLUELINK,
    anchorcolor=DARKBLUELINK,
    citecolor=DARKBLUELINK,
    filecolor=DARKBLUELINK,
    menucolor=DARKBLUELINK,
    urlcolor=BLUELINK
} % Color citation links in purple
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{naturalnames}{hyperref}

\usepackage{biorxiv}
\usepackage[backend=biber,eprint=false,isbn=false,url=false,giveninits=true, style=apa,date=year]{biblatex}
\DeclareDelimFormat{nameyeardelim}{\addcomma\space}
\addbibresource{references.bib}

@@

% -- Defining colors:
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}% Definig a custom style:
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    basicstyle=\ttfamily\scriptsize\bfseries,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=t,
    keepspaces=true,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}% -- Setting up the custom style:
\lstset{style=mystyle}
\captionsetup[table]{hypcap=false}
\captionsetup[figure]{hypcap=false}

\newcommand{\defEqual}{\stackrel{\text{def}}{=}}
\newcommand{\Multiply}{\cdot}
\newcommand{\MultiplyMatrix}{\times}
\newcommand{\UniDimArray}[1]{\bm{#1}}
\newcommand{\BiDimArray}[1]{\bm{#1}}
\newcommand{\tr}{^{\intercal}}
\newcommand{\inv}{^{-1}}
\DeclareMathOperator{\E}{\mathbb{E}}
\DeclareMathOperator{\Var}{\text{Var}}
\DeclareMathOperator{\Cov}{\text{Cov}}
\newcommand{\der}{\mathrm{d}}
\newcommand{\e}{\text{e}}
\newcommand{\Ne}{N_{\text{e}}}
\newcommand{\dn}{d_N}
\newcommand{\ds}{d_S}
\newcommand{\dnds}{\dn / \ds}
\newcommand{\pn}{\pi_N}
\newcommand{\ps}{\pi_S}
\newcommand{\pnps}{\pn / \ps}
\newcommand{\proba}{\mathbb{P}}
\newcommand{\pfix}{\proba_{\text{fix}}}
\newcommand{\Spi}{i}
\newcommand{\Spj}{j}
\newcommand{\NbrTaxa}{n}
\newcommand{\Time}{t}
\newcommand{\Trait}{P}
\newcommand{\Heritability}{h^2}
\newcommand{\HeritabilitySpi}{\Heritability_{\Spi}}
\newcommand{\MeanTrait}{\bar{\Trait_{\Time}}}
\newcommand{\VecTrait}{\UniDimArray{\bar{\Trait}}}
\newcommand{\Root}{0}
\newcommand{\RootTrait}{\Trait_{\Root}}
\newcommand{\VarPhy}{\Var \left[\MeanTrait\right]}
\newcommand{\VecZero}{\UniDimArray{0}}
\newcommand{\VecOne}{\UniDimArray{1}}
\newcommand{\Distance}{\BiDimArray{D}}
\newcommand{\DistanceMatrix}{\BiDimArray{\Distance}}
\newcommand{\MutationRate}{\mu}
\newcommand{\SubRate}{q}
\newcommand{\NbrLoci}{L}
\newcommand{\VarPhenotype}{V_{\Trait}}
\newcommand{\VarPhenotypeSpi}{V_{\Trait, \Spi}}
\newcommand{\VarGenetic}{V_{\mathrm{A}}}
\newcommand{\VarGeneticSpi}{V_{\mathrm{A}, \Spi}}
\newcommand{\VarEnv}{V_{\mathrm{E}}}
\newcommand{\VarMutation}{V_{\mathrm{M}}}
\newcommand{\GenArchi}{\NbrLoci \Multiply \E \left[ a^2 \right]}
\newcommand{\RateMut}{\sigma^2_{\mathrm{M}}}
\newcommand{\RateBetween}{\sigma^2_{\mathrm{B}}}
\newcommand{\RateWhithin}{\sigma^2_{\mathrm{W}}}
\newcommand{\RateWhithinSpi}{\sigma^2_{\mathrm{W}, \Spi}}
\newcommand{\VecRateWhithin}{\UniDimArray{\RateWhithin}}
\newcommand{\EstRateBetween}{\widehat{\RateBetween}}
\newcommand{\EstRateWhithin}{\widehat{\RateWhithin}}
\newcommand{\NI}{\rho}
\newcommand{\EstNI}{\widehat{\rho}}
\newcommand{\StdSelection}{\sigma}
\newcommand{\VarSelection}{\StdSelection^2}

@@

% Including genomics, Inclusion of genomics, Linking genomics,
\title{Leveraging genomics to detect trait selection from within and between-species variations}
\rhead{\scshape Trait selection from within and between-species variations}

\author{
\large
\textbf{T. {Latrille}$^{1}$\orcidlink{0000-0002-9643-4668}, M. {Bastian}$^{2}$, T. {Gaboriau}$^{1}$\orcidlink{0000-0001-7530-2204}, N. {Salamin}$^{1}$\orcidlink{0000-0002-3963-4954}}\\
\normalsize
$^{1}$Department of Computational Biology, Université de Lausanne, Lausanne, Switzerland\\
$^{2}$Laboratoire de Biométrie et Biologie Evolutive, UMR5558, Université Lyon 1, Villeurbanne, France \\
\texttt{\href{mailto:thibault.latrille@ens-lyon.org}{thibault.latrille@ens-lyon.org}} \\
}
\author{~}
\begin{document}

@@

% Abstract (≤ 250 words)
%TC:endignore
\begin{abstract}
    To determine whether a trait is neutrally evolving or under selection, methods have been developed using either within or between-species variation.
    However, methods using within-species variation do not integrate the changes at the macro-evolutionary scale.
    Conversely, current methods using between-species variation usually discard within-species variation, thus not accounting for processes at the micro-evolutionary scale.
    The main goal of this study is to define a neutrality index for a quantitative trait, by combining within- and between-species variation.
    This neutrality index integrates nucleotide polymorphism and divergence for normalizing trait variation.
    As such, it does not require estimation of population size nor of time of speciation for normalization.
    Our index can be used to seek deviation from the null model of neutral evolution, and test for diversifying selection.
    Applied to brain mass and body mass at the mammalian scale, we show that brain mass is under diversifying selection.
    Finally, we show that our test is not sensitive to the assumption that population sizes, mutation rates and generation time are constant across the phylogeny, and automatically adjust for it.
\end{abstract}

\keywords{Quantitative genetics \and Trait evolution \and Selection \and Phylogenetics \and Population genetics }

@@

Determining whether a trait is neutrally evolving or under a particular regime of selection has been a long-standing goal in evolutionary biology.
Fundamentally, distinguishing neutral evolution from selection requires determining which selective regime is supported by the observed variation of traits or sequences.
The variation of phenotypes (traits) and genotypes (sequences) can be observed at different scales, across different development stages at the individual level, across different individuals and populations at the species level, and finally across different species at the phylogenetic level.
All these systems require different assumptions and methodologies, and the endeavor to determine the selective regime for a given trait has thus incorporated theories, methods, and developments across various fields of evolutionary biology such as quantitative genetics, population genetics, phylogenetics and comparative genomics~\parencite{lynch_genetics_1998, walsh_evolution_2018}.

[not sure of this paragraph...]

Leveraging individual variations within the same species, Genome-Wide Association Studies (GWAS) in humans have shown that traits are mostly polygenic (many loci associated with a given trait), are also mostly pleiotropic (many traits associated with a given locus), additive and mainly under stabilizing selection~\parencite{simons_population_2018, sella_thinking_2019}.
Additionally, contrasting both trait and genetic differentiation across several populations, Q$_\text{ST}$--F$_\text{ST}$ methods have allowed to determine the selective regime and to quantify the strength of selection acting on a trait~\parencite{martin_multivariate_2008, leinonen_qst_2013}.
For example, studies have shown that stabilizing selection is largely dominant in the evolution of gene expression in different populations~\parencite{whitehead_neutral_2006, gilad_natural_2006}. [why only example from gene expression?]
However, dominant selective regime acting on gene expression is still controversial, and neutral evolution is still debated~\parencite{signor_evolution_2018, price_detecting_2022}.

To disentangle neutral evolution and selection, trait evolution can also be observed at a larger time scale.
For example, change in mean trait value accumulates linearly with time of divergence from a sister species, and also proportionally to the trait variance~\parencite{lande_genetic_1980, turelli_heritable_1984}.
Gene expression exhibits a similar accumulation as divergence in expression accumulates faster for gene with large within-species variation~\parencite{khaitovich_neutral_2004}.
Altogether, both the trait variance and the evolution in mean value can be used to test for trait selection in a pair of species~\parencite{walsh_evolution_2018}.

% Phylogenetics (mean trait evolution)
Alternatively, by accounting for the underlying relationships between several species, the selective regime for a quantitative trait can also be tested at the phylogenetic scale~\parencite{felsenstein_phylogenies_1985}.
Under neutral evolution, the change in main trait value along a given branch of the tree is normally distributed, with a variance proportional to divergence time~\parencite{hansen_translating_1996}.
As a result, the mean trait value can be modeled as a Brownian process branching at every node of the tree~\parencite{hansen_translating_1996, harmon_phylogenetic_2018}.
Reconstructing the trait variation along the whole phylogeny as a Brownian process can thus constitute a null model of neutral trait evolution.
Some deviations from the Brownian process, typically Ornstein-Uhlenbeck processes, are thus often interpreted as a signature of stabilizing selection~\parencite{catalan_drift_2019}.
Alternatively, a trend in the Brownian process (the tendency of a trait to evolve in a certain direction) is interpreted as a signature of directional selection at the phylogenetic scale~\parencite{silvestro_early_2019}.
However, studies have shown that such comparative approaches are subject to different biases~\parencite{harmon_phylogenetic_2018}.
First, a trait under stabilizing selection for which the optimal trait value is also evolving as a Brownian process will not deviate from a Brownian process, and thus be wrongly classified as neutral~\parencite{hansen_translating_1996}.
In other words, the better fit of a Brownian process does not necessarily constitute proof of the neutral model.
Secondly, and contrarily, even under a neutral trait, the Ornstein-Uhlenbeck process might sometimes be statistically preferred over a Brownian process due to sampling artifacts~\parencite{silvestro_measurement_2015, cooper_cautionary_2016, price_detecting_2022}.
Those limitations, altogether with the use of mean trait estimates leaving out the variance in traits between individuals, easily generate misclassification of selection from methods at the phylogenetic scale.

% At the frontier Phylogenetics/Population-genetics
At the frontier between micro and macroevolution, comparative methods at the phylogenetic scale have acknowledged the importance of modeling within-species variation on top of changes in mean trait value~\parencite{fay_evaluating_2008, kostikova_bridging_2016, gaboriau_multiplatform_2020, gaboriau_exploring_2023}.
However, in the context of detecting selection acting on a trait, within-species variation is mostly used to describe technical errors.
One notable exception used the ratio of between to within variation estimated for many traits, and compared to the average over all traits to seek deviation from this average~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}.
The goal of this study is to determine the evolutionary regime of a quantitative trait through the articulation of the variance between and within species while setting the threshold for neutral, stabilizing, and diversifying selection.
We propose a neutrality index for a quantitative trait articulating trait and nucleotide variation within and between species.
Importantly, our neutrality index also leverages nucleotide divergence and polymorphism in order to normalize trait variation at both scales, such that it does not require estimating population size (within-species) or speciation time (between-species).
For population geneticists, our study can be seen as the macro-evolutionary generalization of Q$_\text{ST}$--F$_\text{ST}$ methods to account for phylogenetic relationships between species.
For phylogeneticists, our study can be seen as an alternative to the EVE model~\parencite{rohlfs_modeling_2014, rohlfs_phylogenetic_2015}, where we set a clear threshold for neutral evolution by reviewing theoretical literature and leveraging species' nucleotide polymorphism and divergence.
% Basically, we are testing whether the variance of means is greater than the mean of variances

\section*{Materials and Methods}\label{sec:materials-and-methods}
\subsection*{Neutrality index for a quantitative trait}\label{subsec:neutrality-index-for-a-quantitative-trait}

Prior to developing our neutrality index, we review theoretical expectations for variations of quantitative traits and genomic sequences under neutral evolution for both within- and between-species variation.

\subsubsection*{Within-species trait variations}

For a given trait, the genetic architecture is mainly defined by the number of loci encoding the trait ($\NbrLoci$) and the random additive effect of a mutation on the trait ($a$).
New mutations are generating trait variance and the average effect of a mutation on the trait is $\RateMut = \NbrLoci \Multiply  \E [a^2]$.
At the individual level, the mutational variance ($\VarMutation$) is the rate at which new mutations contribute to the trait variance per generation.
As shown in \textcite{lande_quantitative_1979, lande_sexual_1980}, $\VarMutation$ is a function of the mutation rate per loci per generation ($\MutationRate$) and $\RateMut$:
\begin{gather}
    \VarMutation = 2 \MutationRate \Multiply \RateMut \label{eq:var-mutation}.
\end{gather}

While in an infinitesimal model mutations supply new genetic variants, random genetic drift depletes standing variation~\parencite{turelli_commentary_2017, barton_infinitesimal_2017, sella_thinking_2019}.
For a neutral trait at equilibrium between mutation and drift~\parencite{lynch_mutation_1998}, the additive genetic variance in a species ($\VarGenetic$) is a function of the mutational variance ($\VarMutation$) and the effective number of individuals in the population ($\Ne$):
\begin{align}
    \VarGenetic & =  2 \Ne \Multiply \VarMutation, \\
    & = 4 \Ne \Multiply \MutationRate \Multiply \RateMut \text{ from eq.~\ref{eq:var-mutation}}\label{eq:var-genetic}.
\end{align}

For any neutral genomic region of interest, the nucleotide diversity, $\pi$, is measured as the number of mutations segregating in the population divided by the length of the region.
Any segregating mutations will eventually reach fixation or extinction due to random genetic drift and $\pi$ is also at a balance between mutations and drift.
As shown in \textcite{tajima_statistical_1989}, $\pi$ is a function of the mutation rate per loci per generation ($\MutationRate$) and the effective population size ($\Ne$):
\begin{gather}
    \pi = 4 \Ne \Multiply \MutationRate \label{eq:genetic-diversity}.
\end{gather}

We define $\RateWhithin$ as the ratio of additive genetic variance of the trait ($\VarGenetic$) over $\pi$ of any neutral genomic region of interest.
This ratio allows to remove the effect of $\Ne$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trait, giving $\RateWhithin$ as a proxy of $\RateMut$:
\begin{align}
    \RateWhithin & \defEqual \frac{\VarGenetic }{\pi}, \\
    & = \frac{4 \Ne \Multiply \MutationRate \Multiply \RateMut}{4 \Ne \Multiply \MutationRate} \text{ from eq.~\ref{eq:var-mutation} and~\ref{eq:genetic-diversity}}, \\
    & = \RateMut. \label{eq:rate-pop}
\end{align}

The additive genetic variance is also equal to the observed phenotypic variance ($\VarPhenotype$) multiplied by narrow-sense heritability ($\Heritability$; \parencite{hill_data_2008}), which leads to $\RateWhithin$ beinf a function of $\VarPhenotype$ and $\Heritability$:
\begin{align}
    \RateWhithin & = \frac{\Heritability \Multiply \VarPhenotype }{\pi }. \label{eq:rate-pheno-pop}
\end{align}

\subsubsection*{Between-species trait variations}

For a given species, we denote by $\MeanTrait$ the mean value of the trait across the individuals in the species at generation $\Time$.
If the trait is neutral and encoded by many loci as assumed by the infinitesimal model, $\MeanTrait$ evolves as a Brownian process~\parencite{felsenstein_phylogenies_1985, hansen_translating_1996}.
The variance of $\MeanTrait$ after $\Time$ generations, $\VarPhy$ is given by \parencite{hansen_translating_1996}:
\begin{align}
    \VarPhy & = \frac{\VarGenetic}{\Ne} \Multiply \Time \\
    & = 4 \Time \Multiply \MutationRate \Multiply \RateMut \label{eq:covar},\text{ from eq.~\ref{eq:var-genetic}},
\end{align}

Moreover, for any genomic region under neutral evolution, some mutations will eventually reach fixation due to random genetic drift, resulting in a substitution of a nucleotide at the species level.
The probability of fixation ($\pfix$) of a neutral mutation is $1/2\Ne$ \parencite{kimura_probability_1962}.
We can derive the substitution rate per generation $\SubRate$ as the number of mutations per generation ($2\Ne \Multiply \MutationRate$) multiplied by the probability of fixation for each newly arisen mutations $\pfix$~\parencite{mccandlish_modeling_2014}, giving:
\begin{align}
    \SubRate & = 2 \Ne \Multiply \MutationRate \Multiply \pfix, \\
    & = 2 \Ne  \Multiply \MutationRate  \Multiply \dfrac{1}{2\Ne}, \\
    & = \MutationRate. \label{eq:substitution-rate}
\end{align}
That is, if mutations are neutral, the rate of substitution within a genomic region equals the rate with which new mutations arise per generation for the same genomic region~\parencite{kimura_evolutionary_1968}.

After $\Time$ generations and assuming that no multiple substitutions occurred at the same site, the nucleotide divergence $d$, which is the fraction of the genomic region that generated a substitution, will be $\Time$ multiplied by the nucleotide substitution rate per generation ($\SubRate$):
\begin{align}
    d & = \Time \Multiply \SubRate \\
    & = \Time \Multiply \MutationRate \text{ from eq.~\ref{eq:substitution-rate}}. \label{eq:distance}
\end{align}
We define $\RateBetween$ as the variance in the mean trait value ($\VarPhy$) normalized by the nucleotide divergence of any neutral genomic region ($d$).
This ratio allows to remove the effect of $\Time$ and $\MutationRate$, which are parameters not related to the genetic architecture of the trade, giving $\RateBetween$ as another proxy of $\RateMut$:
\begin{align}
    \RateBetween & \defEqual \frac{\VarPhy}{4 d},\\
    & = \frac{4 \Time \Multiply \MutationRate \Multiply \RateMut}{4 \Time \Multiply \MutationRate}\text{ from eq.~\ref{eq:covar} and~\ref{eq:distance}}, \\
    & = \RateMut \label{eq:rate-phy}.
\end{align}
